---
title: "MTH6139 Time Series Coursework 1" 
subtitle: "Time Series Analysis on Seatbelts Dataset Using Prophet"
author:
- name: Yusuf Khan
date: "March 2024"
output:
  html_document:
    toc: true
    toc_float: true
---


# Forecasting using Meta's Prophet

## 1.1 : What is prophet?

Prophet from meta is a procedure which forecasts time series data. It can be used in r by `install.packages("prophet")`.

## 1.2 : Seatbelts Data

I will doing a time series analysis using the Seatbelts dataset in R. This is data of drivers in Great Britain killed with monthly time periods and yearly seasonality.  Because we are only interested in total number of drivers killed, we will look only at the first column. In order to use `prophet()` the data needs to be a data frame with the time column called ds and the data column called y. This is essential for prophet to work. We also need the to convert the times into a vector of dates (as R usually sees this as just some random characters). For this we use the zoo package.

```{r}
date <- zoo::as.yearmon(time(Seatbelts))
date
Seatbelts.df <- data.frame(ds = date, y = Seatbelts[,1])
head(Seatbelts.df)
```

## 1.3 : Plot and Decomposition

We can plot the Time Series and We can also use the function `plot(decompose())` to plot the components of the Time Series in order to visualise it. These are the trend, the seasonality and the noise (residual error).

```{r}
plot(Seatbelts[,1])
plot(decompose(Seatbelts[,1]))
```

From the plot of the Time Series we can see that there is no obvious change in variance with time so it seems the Time Series is Homoskedastic.
From the decompose plots, we can see that there is an overall decreasing trend. We can also see a pattern in the seasonality; there is an increase towards  the end of the year. Later on, we will see the same thing using Prophet. From the residual error plot, we can see that there is no obvious trend or change in variance so it seems like it is stationary.

## 1.4 Forecasting

Now that we have formatted the data properly we can use prophet to forecast. We first fit the model using prophet and then we need to make a new data frame of how much time we wish to forecast (predict). We will be forecasting one year and since we have monthly time periods we will do monthly frequency and 12 periods.

```{r}
model <- prophet::prophet(Seatbelts.df)
forecastdates <- prophet::make_future_dataframe(model, periods = 12, freq = "month")
forecast <- predict(model, forecastdates)
plot(model, forecast)
```

The dots on the plot are the historical observed value from data. The blue line is the forecasted (predicted) values and the shaded blue area is the uncertainty of our forecast.

## 1.5 : Decomposition using Prophet

Similar to before, but now also using our forecasted values, we can plot the components of our Time Series using prophet.

```{r}
prophet::prophet_plot_components(model, forecast)
```

From this we can again see that there is an overall decline in the trend; the total number of drivers killed has been decreasing over time. We can also see the trend in seasonality; the trend during the year. It shows that there seems to be an increase in deaths during the winter periods which could be due to icy road conditions.










